[Re] Spike Timing Dependent Plasticity Finds the Start of Repeating Patterns in Continuous Spike Trains

### 2018-05-27 article/Hathway-Goodman-2018.tex
Added final (?) versions of article, readme files, code

\documentclass[10pt,a4paper,onecolumn]{article}
% \usepackage[utf8]{inputenc}
\usepackage{marginnote}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage{authblk,etoolbox}
\usepackage{titlesec}
\usepackage{calc}
\usepackage{hyperref}
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor=
{
      Pamela Hathway,
      Dan Goodman,,
  },
            pdftitle=
{
[Re] Spike Timing Dependent Plasticity Finds the Start of Repeating Patterns
in Continuous Spike Trains
},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=blue,
            pdfborder={0 0 0}}
\urlstyle{same}
\usepackage{tcolorbox}
\usepackage{ragged2e}
\usepackage{fontspec}
\defaultfontfeatures{Path = /usr/local/texlive/2018/texmf-dist/fonts/opentype/public/fontawesome/}
\usepackage{fontawesome}
\usepackage{caption}
\usepackage{listings}
\lstnewenvironment{code}{\lstset{language=Haskell,basicstyle=\small\ttfamily}}{}

@@

    \usepackage{authblk}
                        \author[1]{Pamela Hathway}
                    \author[1]{Dan Goodman,}
                            \affil[1]{Department of Electrical and Electronic Engineering, Imperial College,
London, UK}

@@

Neurons communicate through repeated, specifically timed action
potential sequences (spike patterns) to convey information
\cite{Fellous2004, Prut1998}. Since neuronal activity is noisy and
neurons are likely involved in a multitude of spike patterns of various
lengths and extent, it can be hard to find spike patterns at first
glance. The more neurons are recorded, the more difficult the task
becomes due to the exponential increase of possible combinations of
spikes that could make up a pattern \cite{Buonomano2009}. It is unclear
how neurons in the brain may extract relevant information from such
input. In a 2008 paper, Masquelier and colleagues demonstrated that a
single neuron with spike timing dependent plasticity (STDP) synapses is
able to find the start of a repeating pattern in noisy (artificial) data
\cite{Masq2008}.

\textcite{Masq2008} use a simple feedforward network in which 2000 input
neurons connect to one output neuron via excitatory STDP synapses. Half
of the input neurons spike according to a spike pattern of 50 ms length
for about 25\% of the time. Finding the pattern is additionally made
difficult by jittering the pattern spikes, adding random noise spikes to
all neurons, and a constant population rate as well as no differences in
overall firing rate between neurons.

We replicated their findings using the spiking neural network simulator
Brian \cite{Goodman2009, Stimberg2014}, whereas the the original study
implemented the simulations in Matlab, with the main functions being
computed in C/C++ through mex files. In addition, we examined some of
the implementation parameters and investigated whether they were
essential to the success of the algorithm.

@@

All simulations were performed using the spiking neural network
simulator Brian (Brian2, version 2.0, http://briansimulator.org/). We
attempted to stay as true as possible to the original study. Simulation
parameters were taken from the text and we additionally obtained the
source code for the standard parameter configuration from the authors to
ensure equivalency of the implementations. The source code for deleting
spikes within the pattern (Figure ~\ref{fig:para} E) was not provided.

@@

\hypertarget{tbl:param}{}
\begin{longtable}[]{@{}lrr@{}}
\caption{\label{tbl:param}Parameter variations}\tabularnewline
\toprule
parameters & standard & variations\tabularnewline
\midrule
\endfirsthead
\toprule
parameters & standard & variations\tabularnewline
\midrule
\endhead
w\(_{initial}\) & 0.475 & 0.275, 0.325, 0.375, 0.425\tabularnewline
jitter (sd) {[}ms{]} & 1 & 0, 2, 3, 4, 5, 6\tabularnewline
pattern frequency & 0.25 & 0.05, 0.1, 0.15, 0.5\tabularnewline
prop. neurons in pattern & 0.5 & 0.3, 0.4, 0.6\tabularnewline
spike deletion & 0 & 0.1, 0.2, 0.3\tabularnewline
\bottomrule
\end{longtable}

@@

The absence of a refractory period in the input neurons has the effect
inter spike intervals can be as small as 10\(^{-9}\) s in some cases.
The original implementation calculates the simulation variables (output
neuron voltage etc.) at every input neuron or output neuron spike, so
such small inter spike intervals are not a problem.

Brian on the other hand uses discrete time steps for its simulations and
updates its simulation variables once per time step. In order to convert
the original spike trains into Brian, we had to modify the spike trains
slightly: whenever two spikes from the same neuron happened in the same
time step, we deleted the second spike. At a resolution of 10\(^{-4}\) s
(standard time step) this affected 0.25\% of all spikes and 0.0025\% at
a resolution of 10\(^{-6}\) s. Deleting spikes does not affect the
amount of incoming voltage significantly since the initial firing rate
of the output neuron is the same with or without the close spikes.

@@

The original study models the potential of the output neuron with the
Gerstner's Spike Response model (SRM) \cite{Gerstner2002}, which uses
kernels to calculate the effect of incoming spikes on the postsynaptic
voltage. Brian on the other hand uses differential equations to model
the system parameters and evaluates those equations for each time step.
We converted the kernels of the posystnaptic potential and the spike
afterpotential into the following differential equations:

@@

The values for the parameters can be found in Table ~\ref{tbl:param}.
Equation ~\ref{eq:1} describes the postsynaptic membrane potential,
which is influenced by presynaptic excitatory postsynaptic potentials
(EPSPs, first term) and a negative afterpotential after a spike (second
term). In the event of a presynaptic spike, x is set to 1, which
initiates the voltage increase in the postsynaptic neuron. When the
postsynaptic voltage reaches the threshold, a spike occurs: the voltage
u is set to twice the threshold, all EPSPs are flushed (all x set to 0)
and the negative afterpotential is set into motion (a is set to 1).
Presynaptic spikes had the same effect on the postsynaptic neuron
potential as in the original publication as shown in Figure
~\ref{fig:pot}.

\hypertarget{tbl:simt}{}
\begin{longtable}[]{@{}lrllr@{}}
\caption{\label{tbl:simt}Simulation parameters}\tabularnewline
\toprule
LIF neuron & value & \ldots{}\ldots{}\ldots{}\ldots{}\ldots{} & STDP &
value\tabularnewline
\midrule
\endfirsthead
\toprule
LIF neuron & value & \ldots{}\ldots{}\ldots{}\ldots{}\ldots{} & STDP &
value\tabularnewline
\midrule
\endhead
\(\tau_{m}\) {[}ms{]} & 10 & & \(\tau_{+}\) {[}ms{]} &
16.8\tabularnewline
\(\tau_{s}\) {[}ms{]} & 2.5 & & \(\tau_{-}\) {[}ms{]} &
33.7\tabularnewline
\(\tau_{syn}\) {[}ms{]} & 2.5 & & \(a_{+}\) & \(2^{-5}\)\tabularnewline
T {[}a.u.{]} & 500 & & \(a_{-}\) & 0.85 \(a_{+}\)\tabularnewline
X & \(\frac{\tau_{s}}{\tau_{m}}^{\frac{\tau_{m}}{\tau_{s} -\tau_{m}}}\)
& & \(w_{min}\) & 0\tabularnewline
A & -3 T & & \(w_{max}\) & 1\tabularnewline
\(\Delta x\) & 1 & & - & -\tabularnewline
\(\Delta a\) & 1 & & - & -\tabularnewline
\bottomrule
\end{longtable}

\begin{figure}
\centering
\includegraphics{figures/fig1_potentials.pdf}
\caption{Postsynaptic potential. The postsynaptic potential of the
replication implementation behaves the same as in the original study as.
Both show the same EPSP shapes and the negative afterpotential after a
postsynaptic spike. \textbf{A}) The postsynaptic potential of the
original study was calculated using the equations given in the original
paper. \textbf{B}) The postsynpatic potential in the replication
implementation was calculated from the differential equations specified
in section \textbf{Leaky Integrate and Fire neuron}}\label{fig:pot}
\end{figure}

@@

For synaptic plasticity, the original study uses the reduced nearest
neighbor rule (RNN) cite\{Morrison2008\} - a more restrictive version of
the nearest neighbor (NN) STDP rule. In the standard NN rule, every
spike causes a weight change (with the amount depending on the timing of
the nearest spike), while for RNN a weight change only happens for the
first postsynaptic spike immediately following a presynaptic spike (or
vice versa). This means that for NN, there can be a series of
potentiations or depressions, whereas for rNN potentiations and
depressions strictly alternate (due to neurons being forced to spike at
least once every 50 ms, see \cite{Masq2008}).

A visualisation of three STDP rules is shown in Figure ~\ref{fig:stdp}:
the RNN, the standard NN and the all-to-all (ATA) rule, in which all
pairs of spikes are considered to calculate the weight change. A case is
considered in which the presynaptic neuron (the input neurons in this
study) spikes more often than the postsynaptic neuron (the output neuron
in this study), leading to synaptic weight decreases for ATA and NN, but
no visible net change under RNN (Figure ~\ref{fig:stdp}\textbf{C}).

@@

The ability to find repeating patterns was reproduced in the replication
implementation. We could qualitatively reproduce all results of the
original paper and show the reproduced figures relevant to the model
behaviour: latency development (Figure ~\ref{fig:lat}), pattern
specificity after convergence (Figure ~\ref{fig:conv}), and robustness
(Figure ~\ref{fig:para}). In addition, we comment on some implementation
details that turned out to be relevant to replicating the original study
successfully: simulation time step, learning rule, EPSP shape.

@@

In order to assess the time of pattern finding, we used the same spike
train (standard conditions) as input for both the original code and in
the replication implementation. The latency development looks similar in
both implementations (see Figure ~\ref{fig:lat}). The time until a
stable state arises is longer in the replication implementation (1400
discharges instead of 700 or 20 s instead of 13.5 s). This discrepancy
is due to the introduction of discrete time steps.

\begin{figure}
\centering
\includegraphics{figures/fig3_latency.pdf}
\caption{Latency. Comparing the latency development of the original
study (\textbf{A}) and the replication implementation (\textbf{B}). In
both implementations, the timing of the output spike in relation to the
start of the pattern is random at first, but becomes selective after a
few hundred discharges. From then on the postsynaptic neuron always
spikes only within the pattern (no spikes at latency = 0). The number of
discharges until this happens is smaller in the original paper than in
the replication implementation.}\label{fig:lat}
\end{figure}

At the end of the simulation, the synaptic weights that are maximally
potentiated (close to w\(_{max}\) = 1) belong exclusively to neurons
involved in the pattern, whereas weights from neurons not involved in
the pattern are depressed nearly completely (see Figure
~\ref{fig:conv}). Neurons that spike at the beginning of the pattern are
potentiated, causing the postsynaptic neuron to spike at a low latency.

\begin{figure}
\centering
\includegraphics{figures/fig4_convergence.pdf}
\caption{Weights at converged state. The weights in the converged state
look very similar in the the original study (\textbf{A)}) and the
replication implementation (\textbf{B)}). Weights from neurons not
involved in the pattern are close to 0 (neurons 1000-1999), whereas
weights from some neurons involved in the pattern (neurons 0-999) are
close to 1 (maximum value). The weights of neurons that spike at the
beginning of the pattern are nearly all close to 1 and their added
postsynaptic potentials cause a spike at the beginning of the pattern
(blue rectangle).}\label{fig:conv}
\end{figure}

@@

The replication implementation behaves the same as the original study
when subjected to different amounts of jitter (Figure ~\ref{fig:para}
\textbf{B}), various proportions of pattern neuron (Figure
~\ref{fig:para} \textbf{C}), different initial weights (Figure
~\ref{fig:para} \textbf{D}) and when a percentage of spikes within the
pattern are deleted (Figure ~\ref{fig:para} \textbf{E}).

\begin{figure}
\centering
\includegraphics{figures/fig5_robustness.pdf}
\caption{Robustness. Pattern finding abilities of the system were tested
against different pattern presentation frequencies (\textbf{A}), jitter
(\textbf{B}), number of neurons spiking according to the pattern
(\textbf{C}), initial weights (\textbf{D}) and deletion of spikes in the
pattern (\textbf{E}). For each parameter combination 100 simulations
were run and the number of successful runs are reported as percentage of
success. The replication implementation was run at two different time
steps: 10\(^{-6}\) s (solid black lines) and 10\(^{-4}\) s (dashed black
lines). The results from the original study were calculated using the
source code provided (solid green lines in \textbf{A}, \textbf{B},
\textbf{C}, \textbf{D}) or estimated from the original publication
(dashed green line in \textbf{E}).}\label{fig:para}
\end{figure}

At a high pattern repetition frequency of 0.5 (Figure ~\ref{fig:para}
\textbf{A}), when the pattern is presented every 100 ms for 50 ms, the
performance of the replication differs from the original: at larger time
steps of 10\(^{-4}\) s the replication version does not perform as well
as the original, but shows good results at smaller time steps
(10\(^{-6}\) s).

@@

We noticed that small implementation details can affect the behaviour of
the network significantly. We summarised the relevant details in Table
~\ref{tbl:impl}\}.

@@

In the standard parameter configuration, the time step chosen does not
have an effect on overall pattern finding success, as long as the time
step is shorter or equal to 10\(^{-4}\) s. At larger time steps
(10\(^{-3}\) s), no specificity emerges, instead there is a systematic
potentiation of all synaptic weights leading to very high firing rates
in the output neuron. In contrast, success rates are above 95\% for
10\(^{-4}\) s, 10\(^{-5}\) s, 10\(^{-6}\) s, and 10\(^{-7}\) s.

Although the success rate stays roughly the same, the time until the
pattern is found (no spikes outside of the pattern happen after this
time point) increases with larger time step size. An example can be seen
in Figure ~\ref{fig:lat}: the pattern is found after about 700 output
neuron spikes in the original publication (left) and after about 1400 in
the replication implementation (right) at a time step of 10\(^{-4}\) s
for the same input. At smaller times steps (10\(^{-6}\) s), the pattern
is found after about 700 discharges - the same as in the original
publication (Figure ~\ref{fig:find}).

\hypertarget{tbl:impl}{}
\begin{longtable}[]{@{}lrr@{}}
\caption{\label{tbl:impl}Implementation details}\tabularnewline
\toprule
& works & does not work\tabularnewline
\midrule
\endfirsthead
\toprule
& works & does not work\tabularnewline
\midrule
\endhead
simulation time step & continuous, discrete &\tabularnewline
time step size & \(\leq\) 10\(^{-4}\) s & \textgreater{} 10\(^{-4}\)
s\tabularnewline
learning rule & RNN & ATA, NN\tabularnewline
EPSP shape & kernel & immediate voltage increase\tabularnewline
\bottomrule
\end{longtable}

@@

The spike trains used here involve neurons with a continuously high
firing rate in the input neurons (on average 64 Hz, min 30 Hz, max 90
Hz) and a significantly lower firing rate in the output neuron (63 Hz
initially, 5Hz after reaching specificity). This means that input
neurons fire often in the time between output neuron spikes. Therefore -
with any other than the reduced NN learning rule - the input neurons
will experience a decrease in weight a lot more often (every time an
input neuron spikes) than an increase in weight (every time the
postsynaptic neuron spikes). This leads to a strong overall depression
of the synaptic weights in the first few seconds of the simulation. With
the parameters specified in \textcite{Masq2008} and an ATA or a
conventional NN learning rule, the output neuron stops firing after a
few seconds because the output neuron voltage does not reach the voltage
threshold necessary to evoke an output neuron spike anymore (see Figure
~\ref{fig:meanw}\}). In the case of the standard parameters, the output
neuron stops firing after less than one second. When no output neuron
spikes occur, learning cannot take place and no specificity can emerge.

\begin{figure}
\centering
\includegraphics{figures/fig7_ata.pdf}
\caption{Effect of learning rules on synaptic weights. \textbf{A)} Both
the ATA and the NN rules lead to a very rapid depression of all synapses
leading to silence in the output neuron. \textbf{B)} In contrast, the
average weight of all neurons declines at a much slower rate for the RNN
rule. \textbf{C)} and \textbf{D)} When tweaking the ATA rule (increasing
a\(_{+}\) substantially) it is possible to achieve a behaviour that
resembles pattern finding. For a short amount of time the output neuron
becomes specific to the pattern, but loses this ability again and does
not regain it.}\label{fig:meanw}
\end{figure}

In the case of an ATA learning rule, a behaviour resembling pattern
finding can be evoked under the right circumstances: reducing the
learning rate by a factor of 5 and increasing the value of a\(_{+}\)
(the maximum weight increase) to double the value of a\(_{-}\) (the
maximum weight decrease). This setup works because the large amount of
depression (on every input spike) is counteracted by large potentiation
(due to the increased a\(_{+}\)). In such runs, the output neuron will
reach specificity and trace back through the latency and then starts
firing outside of the pattern again. This system is not stable since the
ATA rule leads to ``too much learning'' as the synaptic weights change
after every single spike. A further reduction in the learning rule will
not results in pattern specificity at all.

The standard NN learning rule, although exhibiting slower learning when
compared to the ATA rule, is unable to accomplish any pattern finding
because the limitation of potentiation to a certain maximum value means
that depression will always outweigh potentiation as it occurs far more
often.

>>>>

\documentclass[10pt,a4paper,onecolumn]{article}
% \usepackage[utf8]{inputenc}
\usepackage{marginnote}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage{authblk,etoolbox}
\usepackage{titlesec}
\usepackage{calc}
\usepackage{hyperref}
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor=
{
      Pamela Hathway,
      Dan F. M. Goodman,,
  },
            pdftitle=
{
[Re] Spike Timing Dependent Plasticity Finds the Start of Repeating Patterns
in Continuous Spike Trains
},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=blue,
            pdfborder={0 0 0}}
\urlstyle{same}
\usepackage{tcolorbox}
\usepackage{ragged2e}
\usepackage{fontspec}
%\defaultfontfeatures{Path = /usr/local/texlive/2018/texmf-dist/fonts/opentype/public/fontawesome/}
\usepackage{fontawesome}
\usepackage{caption}
\usepackage{listings}
\lstnewenvironment{code}{\lstset{language=Haskell,basicstyle=\small\ttfamily}}{}

@@

    \usepackage{authblk}
                        \author[1]{Pamela Hathway}
                    \author[1]{Dan F. M. Goodman,}
                            \affil[1]{Department of Electrical and Electronic Engineering, Imperial College,
London, UK}

@@

Neurons communicate through repeated, specifically timed action
potential sequences (spike patterns) to convey information
\autocites{Fellous2004}{Prut1998}. Since neuronal activity is noisy and
neurons are likely involved in a multitude of spike patterns of various
lengths and extent, it can be hard to find spike patterns at first
glance. The more neurons are recorded, the more difficult the task
becomes due to the exponential increase of possible combinations of
spikes that could make up a pattern \autocite{Buonomano2009}. It is
unclear how neurons in the brain may extract relevant information from
such input. In a 2008 paper, Masquelier and colleagues demonstrated that
a single neuron with afferent synapses exhibiting spike timing dependent
plasticity (STDP) is able to find the start of a repeating pattern in
noisy (artificial) data \autocite{Masq2008}.

\textcite{Masq2008} use a simple feedforward network in which 2000 input
neurons connect to one output neuron via excitatory STDP synapses. Half
of the input neurons spike according to a spike pattern of 50 ms length
for about 25\% of the time. Finding the pattern is made more difficult
by jittering the pattern spikes, adding random noise spikes to all
neurons, and ensuring a constant population rate as well as no
differences in overall firing rate between neurons.

We replicated their findings using the spiking neural network simulator
Brian \autocites{Goodman2009}{Stimberg2014}, whereas the the original
study implemented the simulations in Matlab, with the main functions
being computed in C/C++ through mex files. In addition, we examined some
of the implementation details and parameters and investigated whether
they were essential to the success of the algorithm.

@@

All simulations were performed using the spiking neural network
simulator Brian (Brian2, version 2.0, http://briansimulator.org/). We
attempted to stay as true as possible to the original study. Simulation
parameters were taken from the text and we additionally obtained the
source code for the standard parameter configuration from the authors to
ensure equivalency of the implementations. The source code for deleting
spikes within the pattern (Fig.~\ref{fig:para} E) was not provided.

@@

\hypertarget{tbl:param}{}
\begin{longtable}[]{@{}lrr@{}}
\caption{\label{tbl:param}Parameter variations }\tabularnewline
\toprule
parameters & standard & variations\tabularnewline
\midrule
\endfirsthead
\toprule
parameters & standard & variations\tabularnewline
\midrule
\endhead
w\(_{initial}\) & 0.475 & 0.275, 0.325, 0.375, 0.425\tabularnewline
jitter (sd) {[}ms{]} & 1 & 0, 2, 3, 4, 5, 6\tabularnewline
pattern frequency & 0.25 & 0.05, 0.1, 0.15, 0.5\tabularnewline
prop. neurons in pattern & 0.5 & 0.3, 0.4, 0.6\tabularnewline
spike deletion & 0 & 0.1, 0.2, 0.3\tabularnewline
\bottomrule
\end{longtable}

@@

The absence of a refractory period in the input neurons has the effect
that inter spike intervals can be as small as 10\(^{-9}\) s in some
cases. The original implementation uses an event-based simulation method
in which the simulation variables (output neuron voltage etc.) are
calculated at every input neuron or output neuron spike, and so such
small inter spike intervals are not a problem.

Brian on the other hand uses discrete time steps for its simulations and
updates its simulation variables once per time step. In order to convert
the original spike trains into Brian, we had to modify the spike trains
slightly: whenever two spikes from the same neuron happened in the same
time step, we deleted the second spike. At a resolution of 10\(^{-4}\) s
(standard time step) this affected 0.25\% of all spikes, and at a
resolution of 10\(^{-6}\) s it affected 0.0025\%. Deleting spikes does
not affect the input drive to the neuron significantly since the initial
firing rate of the output neuron is the same with or without the close
spikes.

@@

The original study models the potential of the output neuron with the
Gerstner's Spike Response model (SRM) \autocite{Gerstner2002}, which
uses kernels to calculate the effect of incoming spikes on the
postsynaptic voltage. Brian on the other hand uses differential
equations to model the system parameters and evaluates those equations
for each time step. We converted the kernels of the posystnaptic
potential and the spike afterpotential into the following differential
equations:

@@

The values for the parameters can be found in Tbl.~\ref{tbl:param}.
Eq.~\ref{eq:1} describes the postsynaptic membrane potential, which is
influenced by presynaptic excitatory postsynaptic potentials (EPSPs,
first term) and a negative afterpotential after a spike (second term).
In the event of a presynaptic spike, \(x\) is increased by 1
(\(x\leftarrow x+1\)), which initiates the voltage increase in the
postsynaptic neuron. When the postsynaptic voltage reaches the
threshold, a spike occurs: the voltage \(u\) is set to twice the
threshold (\(u\leftarrow 2T\)), all EPSPs are flushed (all \(x\) set to
0, \(x\leftarrow 0\)) and the negative afterpotential is set into motion
(\(a\) is set to 1, \(a\leftarrow 1\)). Presynaptic spikes had the same
effect on the postsynaptic neuron potential as in the original
publication as shown in Fig.~\ref{fig:pot}.

\hypertarget{tbl:simt}{}
\begin{longtable}[]{@{}lrllr@{}}
\caption{\label{tbl:simt}Simulation parameters }\tabularnewline
\toprule
LIF neuron & value & \ldots{}\ldots{}\ldots{}\ldots{}\ldots{} & STDP &
value\tabularnewline
\midrule
\endfirsthead
\toprule
LIF neuron & value & \ldots{}\ldots{}\ldots{}\ldots{}\ldots{} & STDP &
value\tabularnewline
\midrule
\endhead
\(\tau_{m}\) {[}ms{]} & 10 & & \(\tau_{+}\) {[}ms{]} &
16.8\tabularnewline
\(\tau_{s}\) {[}ms{]} & 2.5 & & \(\tau_{-}\) {[}ms{]} &
33.7\tabularnewline
\(\tau_{syn}\) {[}ms{]} & 2.5 & & \(a_{+}\) & \(2^{-5}\)\tabularnewline
T {[}a.u.{]} & 500 & & \(a_{-}\) & 0.85 \(a_{+}\)\tabularnewline
X & \(\frac{\tau_{s}}{\tau_{m}}^{\frac{\tau_{m}}{\tau_{s} -\tau_{m}}}\)
& & \(w_{min}\) & 0\tabularnewline
A & -3 T & & \(w_{max}\) & 1\tabularnewline
\(\Delta x\) & 1 & & - & -\tabularnewline
\(\Delta a\) & 1 & & - & -\tabularnewline
\bottomrule
\end{longtable}

\begin{figure}
\centering
\includegraphics{figures/fig1_potentials.pdf}
\caption{Postsynaptic potential. The postsynaptic potential of the
replication implementation behaves the same as in the original study.
Both show the same EPSP shapes and the negative afterpotential after a
postsynaptic spike. \textbf{A}) The postsynaptic potential of the
original study was calculated using the equations given in the original
paper. \textbf{B}) The postsynaptic potential in the replication
implementation was calculated from the differential equations specified
in section \textbf{Leaky Integrate and Fire neuron}}\label{fig:pot}
\end{figure}

@@

For synaptic plasticity, the original study uses the reduced nearest
neighbor rule (RNN) \autocite{Morrison2008} - a more restrictive version
of the nearest neighbor (NN) STDP rule. In the standard NN rule, every
spike causes a weight change (with the amount depending on the timing of
the nearest spike), while for RNN a weight change only happens for the
first postsynaptic spike immediately following a presynaptic spike (or
vice versa). This means that for NN, there can be a series of
potentiations or depressions, whereas for RNN potentiations and
depressions strictly alternate.

A visualisation of three STDP rules is shown in Fig.~\ref{fig:stdp}: the
RNN, the standard NN and the all-to-all (ATA) rule, in which all pairs
of spikes are considered to calculate the weight change. A case is
considered in which the presynaptic neuron (the input neurons in this
study) spikes more often than the postsynaptic neuron (the output neuron
in this study), leading to synaptic weight decreases for ATA and NN, but
no visible net change under RNN (Fig.~\ref{fig:stdp} \textbf{C}).

@@

The ability to find repeating patterns was reproduced in the replication
implementation. We could qualitatively reproduce all results of the
original paper and show the reproduced figures relevant to the model
behaviour: latency development (Fig.~\ref{fig:lat}), pattern specificity
after convergence (Fig.~\ref{fig:conv}), and robustness
(Fig.~\ref{fig:para}). In addition, we comment on some implementation
details that turned out to be relevant to replicating the original study
successfully: simulation time step, learning rule, EPSP shape.

@@

In order to assess the time of pattern finding, we used the same spike
train (standard conditions) as input for both the original code and in
the replication implementation. The latency development looks similar in
both implementations (see Fig.~\ref{fig:lat}). The time until a stable
state arises is longer in the replication implementation (1400
discharges instead of 700 or 20 s instead of 13.5 s). This discrepancy
is due to the size of the discrete time steps used (see section
\textbf{Implementation details}).

\begin{figure}
\centering
\includegraphics{figures/fig3_latency.pdf}
\caption{Latency. Comparing the latency development of the original
study (\textbf{A}) and the replication implementation (\textbf{B}). In
both implementations, the timing of the output spike in relation to the
start of the pattern (vertical axis, set to latency=0 if output spike
happens outside the pattern) is random at first, but becomes selective
after a few hundred discharges. From then on the postsynaptic neuron
always spikes only within the pattern (no spikes at latency = 0). The
number of discharges until this happens is smaller in the original paper
than in the replication implementation.}\label{fig:lat}
\end{figure}

At the end of the simulation, the synaptic weights that are maximally
potentiated (close to w\(_{max}\) = 1) belong exclusively to neurons
involved in the pattern, whereas weights from neurons not involved in
the pattern are depressed nearly completely (see Fig.~\ref{fig:conv}).
Neurons that spike at the beginning of the pattern are potentiated,
causing the postsynaptic neuron to spike at a low latency.

\begin{figure}
\centering
\includegraphics{figures/fig4_convergence.pdf}
\caption{Weights at converged state. The weights in the converged state
look very similar in the the original study (\textbf{A)}) and the
replication implementation (\textbf{B)}). Weights from neurons not
involved in the pattern are close to 0 (black, neurons 1000-1999),
whereas weights from some neurons involved in the pattern (neurons
0-999) are close to 1 (white, maximum value). The weights of neurons
that spike at the beginning of the pattern are nearly all close to 1 and
their added postsynaptic potentials cause a spike at the beginning of
the pattern (blue rectangle).}\label{fig:conv}
\end{figure}

@@

The replication implementation behaves the same as the original study
when subjected to different amounts of jitter (Fig.~\ref{fig:para}
\textbf{B}), various proportions of neurons involved in the pattern
(Fig.~\ref{fig:para} \textbf{C}), different initial weights
(Fig.~\ref{fig:para} \textbf{D}) and when a percentage of spikes within
the pattern are deleted (Fig.~\ref{fig:para} \textbf{E}).

\begin{figure}
\centering
\includegraphics{figures/fig5_robustness.pdf}
\caption{Robustness. Pattern finding abilities of the system were tested
against different pattern presentation frequencies (\textbf{A}), jitter
(\textbf{B}), number of neurons spiking according to the pattern
(\textbf{C}), initial weights (\textbf{D}) and deletion of spikes in the
pattern (\textbf{E}). For each parameter combination 100 simulations
were run and the number of successful runs are reported as percentage of
success. The replication implementation was run at two different time
steps: 10\(^{-4}\) s (solid black lines) and 10\(^{-6}\) s (dashed black
lines). The results from the original study were calculated using the
source code provided (solid green lines in \textbf{A}, \textbf{B},
\textbf{C}, \textbf{D}) or estimated from the original publication
(dashed green line in \textbf{E}).}\label{fig:para}
\end{figure}

At a high pattern repetition frequency of 0.5 (Fig.~\ref{fig:para}
\textbf{A}), when the pattern is presented every 100 ms for 50 ms, the
performance of the replication differs from the original: at larger time
steps of 10\(^{-4}\) s the replication version does not perform as well
as the original, but shows good results at smaller time steps
(10\(^{-6}\) s).

@@

We noticed that small implementation details can affect the behaviour of
the network significantly. We summarised the relevant details in Table
tbl.~\ref{tbl:impl}.

@@

In the standard parameter configuration, the time step chosen does not
have an effect on overall pattern finding success, as long as the time
step is less than or equal to 10\(^{-4}\) s. At larger time steps
(10\(^{-3}\) s), no specificity emerges. Instead there is a systematic
potentiation of all synaptic weights leading to very high firing rates
in the output neuron. In contrast, success rates are above 95\% for
10\(^{-4}\) s, 10\(^{-5}\) s, 10\(^{-6}\) s, and 10\(^{-7}\) s.

Although the success rate stays roughly the same, the time until the
pattern is found (defined as the time after which no output spikes
happen outside of the pattern) increases with larger time step size. An
example can be seen in Fig.~\ref{fig:lat}: the pattern is found after
about 700 output neuron spikes in the original publication (left) and
after about 1400 in the replication implementation (right) at a time
step of 10\(^{-4}\) s for the same input. At smaller times steps
(10\(^{-6}\) s), the pattern is found after about 700 discharges - the
same as in the original publication (Fig.~\ref{fig:find}).

\hypertarget{tbl:impl}{}
\begin{longtable}[]{@{}lrr@{}}
\caption{\label{tbl:impl}Implementation details }\tabularnewline
\toprule
& works & does not work\tabularnewline
\midrule
\endfirsthead
\toprule
& works & does not work\tabularnewline
\midrule
\endhead
simulation time step & continuous, discrete &\tabularnewline
time step size & \(\leq\) 10\(^{-4}\) s & \textgreater{} 10\(^{-4}\)
s\tabularnewline
learning rule & RNN & ATA, NN\tabularnewline
EPSP shape & kernel & immediate voltage increase\tabularnewline
\bottomrule
\end{longtable}

@@

The spike trains used here involve neurons with a continuously high
firing rate in the input neurons (on average 64 Hz, min 30 Hz, max 90
Hz) and a significantly lower firing rate in the output neuron (63 Hz
initially, 5Hz after reaching specificity). This means that input
neurons fire often in the time between output neuron spikes. Therefore,
with any other than the reduced NN learning rule, the input neurons will
experience a decrease in weight a lot more often (every time an input
neuron spikes) than an increase in weight (every time the postsynaptic
neuron spikes). This leads to a strong overall depression of the
synaptic weights in the first few seconds of the simulation. With the
parameters specified in \textcite{Masq2008} and an ATA or a conventional
NN learning rule, the output neuron stops firing after a few seconds
because the output neuron voltage does not reach the voltage threshold
necessary to evoke an output neuron spike anymore (see
Fig.~\ref{fig:meanw}). In the case of the standard parameters, the
output neuron stops firing after less than one second. When no output
neuron spikes occur, learning cannot take place and no specificity can
emerge.

\begin{figure}
\centering
\includegraphics{figures/fig7_ata.pdf}
\caption{Effect of learning rules on synaptic weights. \textbf{A)} Both
the ATA and the NN rules lead to a very rapid depression of all synapses
leading to silence in the output neuron. \textbf{B)} In contrast, the
average weight of all neurons declines at a much slower rate for the RNN
rule. \textbf{C)} and \textbf{D)} When tweaking the ATA rule (increasing
\(a_{+}\) substantially) it is possible to achieve a behaviour that
resembles pattern finding. For a short amount of time the output neuron
becomes specific to the pattern, but loses this ability again and does
not regain it.}\label{fig:meanw}
\end{figure}

In the case of an ATA learning rule, a behaviour resembling pattern
finding can be evoked under the right circumstances: reducing the
learning rate by a factor of 5 and increasing the value of \(a_{+}\)
(the maximum weight increase) to double the value of \(a_{-}\) (the
maximum weight decrease). This setup works because the large amount of
depression (on every input spike) is counteracted by large potentiation
(due to the increased \(a_{+}\)). In such runs, the output neuron will
correctly reach specificity and trace back through the latency, but then
starts firing outside of the pattern again. This system is not stable
since the ATA rule leads to ``too much learning'' as the synaptic
weights change after every single spike. A further reduction in the
learning rule will not result in pattern specificity at all.

We were unable to find parameters for the standard NN learning rule that
allowed for pattern finding, despite this rule exhibiting slower
learning when compared to the ATA rule. In the case of both the ATA and
NN learning rules, it seems likely that stable pattern finding relies on
a precise balance of \(a_{+}\) and \(a_{-}\), with runaway potentiation
or depression likely if the balance is wrong. By contrast, the RNN rule
is automatically balanced and is not subject to this issue.

\subsubsection{\texorpdfstring{Effect of learning rule at
\(\Delta t=0\)}{Effect of learning rule at \textbackslash{}Delta t=0}}\label{effect-of-learning-rule-at-delta-t0}

The timing of the output neuron spikes is slightly different in the
original study and the replication. In the original study spike times of
the input and output neurons are not restricted to fixed multiples of
the timestep, so it is extremely unlikely that two neurons will spike at
the same time. In Brian, the output neuron spikes at the beginning of a
time step and will therefore happen in the same time bin as some input
neuron spikes leading to a time difference between the spikes of
\(\Delta t=0\) where the STDP rule is undefined. Brian treats these
input neuron spikes as if they happened just before the output neuron
spike (\(\Delta t<0\), due to the scheduling of events in Brian) and
will therefore increase all those weights instead of increasing some and
decreasing others. This higher number of potentiations makes it more
difficult for the system to systematically depress unimportant weights
in order to become selective to the pattern. If the learning rule is
modified so that the synaptic weight is increased as well as depressed
when input neuron and output neuron spikes happen during the same time
step (effectively causing a smaller weight increase), the pattern is
found earlier, at around 17 s or 850 spikes (for a time step of
\(10^{-4}s\)) which is close to the performance for smaller time steps
(\(10^{-6}s\)) and the original paper (both 14 s or 700 spikes). We
didn't use this modified learning rule in the rest of the paper because
it differs from the implementation used in the majority of modelling
papers.
